A Switching Planner for Combined Task and Observation Planning

### 2010-11-27 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/dtpddl.tex
spellers

Because PPDDL cannot model domains that feature partial observability,
we developed an extension of that language, called Decision-Theoretic
(DT)PDDL. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~\prob_1 (T_1) .. \prob_n
(T_n))$ where $T_i$ are conjunctive, or (3) a conjunct over
probabilistic and atomic terms. The root term is always conjunctive,
and the leaves are atomic. For example, the starting distribution for
$\pp{DORA}$ might be described thusly:\footnote{In PDDL,
$(\pp{:init}~T_1..T_n)$ expresses the conjunctive root of the tree --
i.e., the root node $(\pp{and}~T_1..T_n)$. Also, we shall write
$\prop$, rather than $(\pp{and}~\prop)$, it a conjunctive term
contains a single atomic subterm.}

@@

\noindent The interpretation of such an expression can be given
according to a {\em visitation} of terms. An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $\prob_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

>>>>

Because PPDDL cannot model domains that feature partial observability,
we develop an extension of that language, called Decision-Theoretic
(DT)PDDL. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~\prob_1 (T_1) .. \prob_n
(T_n))$ where $T_i$ are conjunctive, or (3) a conjunct over
probabilistic and atomic terms. The root term is always conjunctive,
and the leaves are atomic. For example, the starting distribution for
$\pp{DORA}$ might be described as follows:\footnote{In PDDL,
$(\pp{:init}~T_1..T_n)$ expresses the conjunctive root of the tree --
i.e., the root node $(\pp{and}~T_1..T_n)$. Also, we shall write
$\prop$, rather than $(\pp{and}~\prop)$, for conjunctive terms
that contain a single atomic subterm.}

@@

\noindent The interpretation of such an expression can be given
according to a {\em visitation} of terms. An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability that it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $\prob_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

### 2010-11-27 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/intro.tex
spellers

Planning for agents acting in partially observable environments with
stochastic actions is extremely computationally
expensive \cite{mdp-complexity}. Even relatively small partially
observable Markov decision problems (POMDPs), represented by tens of
propositional variables, produce state spaces far larger than existing
optimal or close to optimal planners can handle. For the easier
problem of planning in completely observable stochastic domains, an
alternative approach has been to use classical sequential planners and
replan when actions have unplanned-for
outcomes \cite{yoon:etal:2007}. However, these approaches aren't so
useful in partially observable domains as replanning is triggered when
the system ends up in a state other than the desired one, and this
state can't in general be determined.

@@

Inspired by the replanning approaches used successfully for MDP
planning \cite{yoon:etal:2007,yoon:etal:2008} we propose an approach
that uses a replanning sequential planner to build the majority of the
plan, but uses a decision-theoretic planner on small subproblems where
reasoning about observations is required to find good plans. We refer
to the complete system as a {\em switching planner}. The approach is
not optimal, particularly as it is based on a satisficing sequential
planner, but it performs better than the sequential planner alone and
is fast enough to be used onboard the robot for decision-making. As an
illustration, consider a robot searching for a box that it knows is
often found in offices. The sequential planner might produce a plan
that assumes that {\sc Room a} is an office, that the box is in {\sc
Room a}, and that the box is in location 1 in {\sc Room a}, and then
goes to {\sc Room a}, goes to location 1 and searches for the
box. When the plan gets executed and the search begins, the
decision-theoretic planner is called to conduct the search, and at
this point may retract the last assumption and build a plan to search
the whole of {\sc Room a} for the box. If this plan fails, it will do
so by disproving one of the assumptions (for example, that {\sc Room
a} is an office), which will then lead to replanning by the sequential
planner based on the newly discovered knowledge.

@@

The switching planner approach gives us some important advantages in
our robotic domain. Firstly, the main source of uncertainty in the
domain is unreliable observations, rather than stochasticity in the
action effects, and this approach is particularly suited to problems
that combine a deterministic task planning problem with observations
as the decision-theoretic planner is only invoked to determine the
values of variables. Secondly, the replanning makes the system very
robust to changing goals and the discovery of new facts about the
world, both of which are features of our domain.

@@

An outline of the paper is as follows. We begin by describing POMDPs with a
propositionally factored representation of states and observations,
and therelationship between this and a flat POMDP representation. Next
we describe our DTPDDL language using a running example of the robot
looking for a box. Then we describe the switching planner, including
how the sequential domain is produced and a plan is generated, and how
this then provides the input for the decision-theoretic
planner. Finally we provide an evaluation of the system on the example
robot domain, and finish with a short survey of related work and some
concluding remarks.

>>>>

Planning for agents acting in partially observable environments with
stochastic actions is extremely computationally
expensive \cite{mdp-complexity}. Even relatively small partially
observable Markov decision problems (POMDPs), represented by tens of
propositional variables, produce state spaces far larger than existing
optimal or close to optimal planners can handle. For the easier
problem of planning in completely observable stochastic domains, an
alternative approach has been to use classical sequential planners and
replan when actions have unplanned-for
outcomes \cite{yoon:etal:2007}. However, these approaches are not as
useful in partially observable domains as replanning is triggered when
the system ends up in a state other than the desired one, and this
state cannot in general be determined.

@@

Inspired by the replanning approaches used successfully for MDP
planning \cite{yoon:etal:2007,yoon:etal:2008} we propose an approach
that uses a replanning sequential planner to build the majority of the
plan, but uses a decision-theoretic planner on small subproblems where
reasoning about observations is required to find good plans. We refer
to the complete system as a {\em switching planner}. The approach is
not optimal, particularly as it is based on a satisficing sequential
planner, but it performs better than the sequential planner alone and
is fast enough to be used on the robot for decision-making. As an
illustration, consider a robot searching for a box that it knows is
often found in offices. The sequential planner might produce a plan
that assumes that {\sc Room a} is an office, that the box is in {\sc
Room a}, and that the box is in location 1 in {\sc Room a}, and then
goes to {\sc Room a}, goes to location 1 and searches for the
box. When the plan gets executed and the search begins, the
decision-theoretic planner is called to conduct the search, and at
this point may retract the last assumption and build a plan to search
the whole of {\sc Room a} for the box. If this plan fails, it will do
so by disproving one of the assumptions (for example, that {\sc Room
a} is an office), which will then lead to replanning by the sequential
planner based on the newly discovered knowledge.

@@

The switching planner approach gives us some important advantages in
our robotic domain. First, the main source of uncertainty in the
domain is unreliable observations, rather than stochasticity in the
action effects, and this approach is particularly suited to problems
that combine a deterministic task planning problem with observations
as the decision-theoretic planner is only invoked to determine the
values of variables. Secondly, the replanning makes the system very
robust to changing goals and the discovery of new facts about the
world, both of which are features of our domain.

@@

An outline of the paper is as follows. We begin by describing POMDPs with a
propositionally factored representation of states and observations,
and the relationship between this and a flat POMDP representation. Next
we describe our DTPDDL language using a running example of the robot
looking for a box. Then we describe the switching planner, including
how the sequential domain is produced and a plan is generated, and how
this then provides the input for the decision-theoretic
planner. Finally we provide an evaluation of the system on the example
robot domain, and finish with a short survey of related work and some
concluding remarks.

### 2010-11-27 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/switching.tex
spellers

Our approach always begins with a sequential session. Non-assumptive
actions from the trace are executed in sequence until the
applicability of the next scheduled action is uncertain. We denote
that action \switchAction.  A contingent session then begins, which
tailors sensory processing by the agent to determine whether the
assumptions made in the trace hold. For each assumption we add a {\em
disconfirm} action to the POMDP whose execution is rewarding
(resp. costly) if the assumption is false (resp. true). We also add
one {\em confirm} action that is rewarding (resp. costly) to execute
if $\poss(\switchAction)$ is true (resp. false). If execution of the
contingent plan applies a disconfirm action, then a new sequential
session begins at the {\em underlying} belief-state. If the confirm
action is executed, the sequential session is resumed, and
\switchAction\ is executed. Otherwise, control rests in the continual
session.
%%
Because contingent planning is only practical in relatively small
POMDPs, contingent sessions plan in an abstract decision process
determined by the current trace. This abstraction is constructed by
first omitting all propositions that do not featured in the trace, and
by then iteratively refining the model while the result is of a
practicable size.

@@

\noindent Where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we omit
assumptions that are not {\em active} with respect to the action,
\switchAction, whose scheduled execution switched the system into
the contingent session. An assumption $\assumptiveS{i}$ is active if
\switchAction\ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix
to \switchAction. For example, if $\pp{DORA}$'s trace is:

@@

For the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\switchAction)$, returns control to a sequential
session. If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\switchAction)$ is
executed, then the current sequential plan is executed until further
sensing is scheduled, or to completion. Lastly, we note that in our
project the rewards, i.e., function $\$:2^\propositions\to\Re$, for
the POMDP posed to the contingent session are sourced from a
motivational component of our robotic architecture. We make no
assumptions about how they are derived.

>>>>

Our approach always begins with a sequential session. Non-assumptive
actions from the trace are executed in sequence until the
applicability of the next scheduled action is uncertain. We denote
that action \switchAction.  A contingent session then begins, that
tailors sensory processing by the agent to determine whether the
assumptions made in the trace hold. For each assumption we add a {\em
disconfirm} action to the POMDP whose execution is rewarding
(resp. costly) if the assumption is false (resp. true). We also add
one {\em confirm} action that is rewarding (resp. costly) to execute
if $\poss(\switchAction)$ is true (resp. false). If execution of the
contingent plan applies a disconfirm action, then a new sequential
session begins at the {\em underlying} belief-state. If the confirm
action is executed, the sequential session is resumed, and
\switchAction\ is executed. Otherwise, control rests in the continual
session.
%%
Because contingent planning is only practical in relatively small
POMDPs, contingent sessions plan in an abstract decision process
determined by the current trace. This abstraction is constructed by
first omitting all propositions that do not featured in the trace, and
by then iteratively refining the model while the result is of a
practicable size.

@@

\noindent where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we omit
assumptions that are not {\em active} with respect to the action,
\switchAction, whose scheduled execution switched the system into
the contingent session. An assumption $\assumptiveS{i}$ is active if
\switchAction\ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix
to \switchAction. For example, if $\pp{DORA}$'s trace is:

@@

For the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\switchAction)$, returns control to a sequential
session. If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\switchAction)$ is
executed, then the current sequential plan is executed until further
sensing is scheduled, or to completion. Lastly, we note that in our
project the rewards, i.e., function $\$:2^\propositions\to\Re$, for
the POMDP posed to the contingent session are sourced from a
motivational component of our robotic architecture. We make no
assumptions here about how they are derived.
